Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-06-06 paper.Rnw
edits through 4.2

We perform six applications of LIME with different tuning parameters to the random forest predictions from all observations in the sine data test set. To do this, we use the R package \emph{limeaid} \footnote{Available on GitHub at https://github.com/goodekat/limeaid} (\Sexpr{packageVersion("limeaid")}), which applies a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (\Sexpr{packageVersion("lime")}) \citep{pedersen:2020}. The forked version allows us to export internal values from \emph{lime}. The tuning parameters are as follows. For each application, we specify that LIME returns the top two features in the explanation, because we expect $x_1$ and $x_2$ to be the important features. Note that the complexity of an explanation goes hand in hand with the number of features selected to be included in the explanation. Setting the number of features to a small number will make the explanation simpler but not necessarily more correct. A quantile bin based simulation method is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. The sixth application of LIME uses a kernel density simulation method. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications.

@@

\begin{figure*}[!thp]
<<sine-exp-scatter, out.width = '\\textwidth', fig.width = 10, fig.height = 4, warning = FALSE>>=

@@

The \data \ explanation only includes two features. In situations were more than two features are included in a LIME explanation, the explanation scatterplots can be extended to a generalized pairs plot \citep{emerson:2013} that includes all pairwise combinations of features. Generalized pairs plots (and scatterplot matrices in general) fail to provide helpful information when the number of features becomes large. It is common for machine learning models to have a large number of features, and therefore, a generalized pairs plot of explanation scatterplots for all features would be ineffective. However, when applying LIME, the user selects the number of features to return in the explanation (as previously mentioned). In the \emph{lime} R package, \citet{pedersen:2020} encourage users to select less than 10 features to include in the explanation. As a result, specifying a small number of features to include in a LIME explanation makes it feasible to use generalized pairs plot of explanation scatterplots.

@@

<<sine-feat-heat, out.width = '0.45\\textwidth', fig.width = 4, fig.height = 4>>=
feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "Case")
@
\caption{\emph{Feature Heatmap}. An example feature heatmap of the explanations from the applications of LIME with different tuning parameters to the \data. The cases from the test set are plotted on the y-axis, and the tuning parameters (simulation methods in this example) are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and tuning parameter. The top facet includes the first features selected by LIME, and the second most important features are included in the bottom facet. The column facets separate the kernel density method from the quantile bin methods. The vertical striping seen in the plot indicates that the LIME explanations are not consistent across tuning parameters.}
\label{fig:sine-feat-heat}
\end{figure}

@@

In current practice, forensic firearm  examiners evaluate whether two bullets come from the same source (are fired from the same gun) or from different sources based on microscopic comparison of the striation patterns engraved on bullets during the firing process (see \autoref{fig:bullet}). This process is based on a visual and therefore subjective assessment of the evidence under a comparison microscope. The lack of objective evaluation and the associated absence of established error rates has first been criticized by the National Research Council \cite{nrc:2009} and later by the President's Council of Advisors on Science and Technology \cite{pcast:2016}.

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas \citet{hare:2016} obtain signatures of  striations from two bullet lands (\autoref{fig:signatures}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest model to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures).

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{./figure-static/bullets.png}
\caption{(Top left) Traditionally rifled gun barrel. Grooves and lands alternate to give bullets a spin during the firing process.
Barrel create markings (striations) on a bullet when fired.
(Top right) Image of a fired bullet. The vertical stripes along the lower half of the bullet show groove and land engraved areas. The land engraved areas contain the microscopic striations created when the bullet passed through the barrel of the gun. (Bottom) Zoomed in image of a land engraved area showing striations (vertical lines).}
\label{fig:bullet}

\includegraphics[width=3in]{./figure-static/signatures.png}
\label{fig:signatures}
\caption{Bullet signatures extracted from two bullets fired through the same barrel. The two signatures come from the same barrel-land and therefore have very similar patterns. The features used in the random forest from \citet{hare:2016} are different metrics that measure the similarity between two such signatures.}
\end{figure}

@@

bullet_features_ordered <- bullet_features_ordered %>%
  mutate(feature = factor(feature, levels = bullet_features_ordered$feature))

@@

The random forest model trained in \citet{hare:2016} is available as object rtrees from the \emph{bulletxtrctr} R package. rtrees is trained on a set of scans obtained from one set of bullets of the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009} and based on \Sexpr{dim(bullet_train)[1]} land-to-land comparisons.

@@

@
\caption{Parallel coordinate plots of rtrees training and testing data. The training data are included in the plots on the top row, and the testing data are included in the plots on the bottom row. The left column contains the observations which are know to be comparisons of lands that are not from the same gun, and the known same source comparisons are on the right. The y-axis shows the standardized feature values, and the x-axis shows the features used to fit the random forest (ordered by feature importance). Each line corresponds to a comparison, and the color of the line represents the associated random forest probability. These figures show a clear relationship between the feature values and the probability returned by the random forest.}
\label{fig:bullet-plot-rfpcp}
\end{figure*}

rtrees was tested on another set of bullets from the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. \autoref{fig:bullet-plot-rfpcp} shows parallel coordinate plots of the rtrees training (top row) and testing (bottom row) datasets. The plots on the left contain comparisons know to be from from different sources, and the plots on the right contain comparisons know to be from the same source. Each line represents one observation in the data, and the color of the line represents the by the corresponding random forest probability. The standardized feature values are plotted on the y-axis, and the features are plotted on the x-axis, which are ordered from left to right from the highest to lowest random forest feature importance.

The majority of observations with random forest probabilities close to 1 have a clear pattern of corresponding feature values as can be seen in the plots where same source is known to be true. The observations where the random forest is wrong in returning high probabilities when same source is known to be true (i.e. same source is true but the random forest probabilities are between 0 and 0.5) have feature values that reflect those of observations where same source is known to be false as seen in the plots on the left. These plots provide insight into why the random forest is producing poor random forest probabilities for these observations.

Since firearm identification is commonly used as evidence for convictions in court cases, it is important to be able to understand and assess the model that is used to quantify the probability that a bullet was fired from a gun. An application of LIME to rtrees could provide an understanding of the key variables used by the random forest model to make a prediction. However, just as it is important to assess the random forest model for this high-stakes application, it is also important to assess the LIME explanations to make sure they are providing a trustworthiness understanding of the random forest model. We will apply LIME to rtrees and use our visual diagnostics to assess different LIME tuning parameters.

@@

@@

@
\caption{Feature heatmap of the 36 LIME applications to the bullet comparison data test set. In addition to faceting the results by simulation method and order a feature was selected by LIME as done in \autoref{fig:sine-feature-heatmap}, this plot has been faceted by the Gower power and whether the the observation is from a match (TRUE) or non-match (FALSE). This plot shows several key insights about the LIME explanations. First, the features selected by LIME for the density simulation methods are mostly the same for all tuning parameters and all cases. Second, the features selected by LIME for the bin based methods are affected by the number of bins but not the Gower power. Third, the explanations produced by LIME using bin based methods result in different important features for the matches and non-matches. \kgc{***Need to rotate the facet labels, so I can make the text bigger.}}
\label{fig:bullet-feature-heatmap}
\end{figure*}

We apply LIME to each prediction from the bullet test data obtained from the 'rtrees' random forest model for each of the four sampling methods (equally spaced bins, quantile bins, kernel density estimation, and normal approximation). For each of the bin based sampling methods, LIME was applied for 2 to 6 bins. It was decided to use 6 bins as the maximum, since the larger the number of bins, the more complex the explanation becomes. Each simulation method was implemented three times with Gower exponents of 0.5, 1, and 10. Thus, a total of $12\times 3=36$ different applications of LIME were performed. These applications of LIME were performed using the R package \emph{limeaid}. Each application was set to return 3 features by LIME and feature selection was performed using the highest weights method.

To get an overview of the LIME explanations from the 36 applications, we consider our heatmap diagnostic plot for comparing the LIME applications first shown in \autoref{fig:bullet-feature-heatmap}. The plot facets the features selected in the LIME explanations by the simulation method, the Gower power, the order the feature was chosen, and whether the observation is from a known match or non-match of bullet signatures. This plot highlights several key features of the LIME explanations from the bullet matching dataset.

\begin{figure*}[!thp]
<<bullet-assess-metric, out.width = '\\textwidth', warning = FALSE, fig.width = 10, fig.height = 4>>=

>>>>

We perform six applications of LIME with different tuning parameters to the random forest predictions from all observations in the sine data test set. To do this, we use the R package \emph{limeaid} \footnote{Available on GitHub at https://github.com/goodekat/limeaid} (\Sexpr{packageVersion("limeaid")}), which applies a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (\Sexpr{packageVersion("lime")}) \citep{pedersen:2020}. The forked version allows us to export internal values from \emph{lime}. The tuning parameters are as follows. For each application, we specify that LIME returns the top two features in the explanation, because we expect $x_1$ and $x_2$ to be the important features. Note that the complexity of an explanation goes hand in hand with the number of features selected to be included in the explanation. Setting the number of features to a small number will make the explanation simpler but not necessarily more correct. A quantile bin based simulation method is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. We use 6 bins as the maximum, since the larger the number of bins, the more complex the explanation becomes. The sixth application of LIME uses a kernel density simulation method. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications.

@@

\begin{figure*}[!thp]
<<sine-exp-scatter, out.width = '\\textwidth', fig.width = 10, fig.height = 3.9, warning = FALSE>>=

@@

The \data \ explanation only includes two features. In situations where more than two features are included in a LIME explanation, the explanation scatterplots can be extended to a generalized pairs plot \citep{emerson:2013} that includes all pairwise combinations of features. Generalized pairs plots (and scatterplot matrices in general) have diminishing value as the number of features increase \citep{jensen:2011} \citep{sweller:2011}. Machine learning models are commonly fit using a large number of features, and therefore, a generalized pairs plot of explanation scatterplots for all features would be ineffective. However, when applying LIME, the user selects the number of features to return in the explanation. In the \emph{lime} R package, \citet{pedersen:2020} encourage users to select less than 10 features to include in the explanation. As a result, specifying a small number of features to include in a LIME explanation makes it feasible to use generalized pairs plot of explanation scatterplots.

@@

<<sine-feat-heat, out.width = '0.45\\textwidth', fig.width = 4, fig.height = 4>>=
feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "Case")
@
\caption{\emph{Feature Heatmap}. An example feature heatmap of the explanations from the applications of LIME with different tuning parameters to the \data. The cases from the test set are plotted on the y-axis, and the tuning parameter values (simulation methods here) are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and tuning parameter value. The plot is faceted by the first features selected by LIME (top facet) and the second most important features (bottom facet). The column facets separate the kernel density method from the quantile bin methods. The vertical striping indicates that the LIME explanations are not consistent across tuning parameters.}
\label{fig:sine-feat-heat}
\end{figure}

@@

@@

# Create an ordered version of the data
bullet_features_ordered <- bullet_features_ordered %>%
  mutate(feature = factor(feature, levels = bullet_features_ordered$feature))

@@

In current practice, forensic firearm examiners evaluate whether two bullets are from the same source (fired from the same gun) or from different sources based on microscopic comparison of the striation patterns engraved on bullets during the firing process (see \autoref{fig:bullet}). The process is based on a visual and therefore subjective assessment of the evidence. The lack of objective evaluation and the associated absence of established error rates has first been criticized by the National Research Council \cite{nrc:2009} and later by the President's Council of Advisors on Science and Technology \cite{pcast:2016}.

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:signatures}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest model (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes \Sexpr{dim(bullet_train)[1]} land-to-land comparisons.

\begin{figure}[!t]
\centering
\includegraphics[width=2.25in]{./figure-static/bullets.png}
\caption{(Top left) Traditionally rifled gun barrel. The grooves and lands alternate to give bullets a spin during the firing process, which create markings (striations) on a bullet when fired. (Top right) Image of a fired bullet. The vertical stripes along the lower half of the bullet show groove and land engraved areas. The land engraved areas contain the microscopic striations created when the bullet passed through the barrel of the gun. (Bottom) Close up of a land engraved area showing striations (vertical lines).}
\label{fig:bullet}

\includegraphics[width=3.25in]{./figure-static/signatures.png}
\label{fig:signatures}
\caption{Example bullet signatures. The bullet signatures are from the same barrel-land and therefore have very similar patterns. The \citet{hare:2016} random forest uses features that measure the similarity between two such signatures.}
\end{figure}

@@

@
\caption{Parallel coordinate plots of the \citet{hare:2016} random forest predictions from the training (top facet) and testing (bottom facet) data. The observations are separated by same source (right column) and different source (left column) The y-axis shows the standardized feature values, and the x-axis shows the features used to fit the random forest (ordered by random forest impurity based feature importance). Each line corresponds to a comparison, and the line color represents the associated random forest probability. There are clear relationships between the feature values and the random forest probabilities.}
\label{fig:bullet-plot-rfpcp}
\end{figure*}

Since firearm identification is commonly used as evidence for convictions in court cases, it is important to be able to understand and assess a model used to quantify the probability that a bullet was fired from a gun. LIME explanations would provide a local explanation for an individual prediction, but just as it is important to assess the model for this high-stakes application, it is also important to assess the LIME explanations. We will demonstrate an assessment of LIME explanations using the visual diagnostics introduced in this paper.

We apply the random forest model from \citet{hare:2016} to another set of bullets from the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. We first consider a global visualization of the relationship between the random forest predictions and the model features by creating a parallel coordinate plot of the training data (top facet) and testing data (bottom facets) predictions (\autoref{fig:bullet-plot-rfpcp}). The majority of same source observations with random forest probabilities close to 1 have a clear pattern of corresponding feature values. The known same source observations where the random forest is wrong (does not return probabilities close to 1) have feature values that reflect those of observations where same source is known to be false.

LIME is implemented using the R package \emph{limeaid} to apply LIME multiple times to all test set observations using different tuning parameters: 12 sampling methods (2-6 equally spaced bins, 2-6 quantile bins, kernel density estimation, and normal approximation) and 3 Gower exponents (0.5, 1, and 10). Thus, a total of $12\times 3=36$ different applications of LIME were performed. We specify that each LIME explanation return 3 features and feature selection was performed using the highest weights method (default option).

@@

To get an overview of the LIME explanations from the 36 applications, we consider a feature heatmap (\autoref{fig:bullet-feature-heatmap}). The plot is faceted vertically by simulation method and the Gower power and horizontally by the order the feature was chosen and whether the observation is a known same source or different source comparison. \kgc{This is where I left off...} This plot highlights several key features of the LIME explanations from the bullet matching dataset.

@@

@
\caption{Feature heatmap of the 36 LIME applications to the bullet comparison data test set. In addition to faceting the results by simulation method and order a feature was selected by LIME as done in \autoref{fig:sine-feature-heatmap}, this plot has been faceted by the Gower power and whether the the observation is from a match (TRUE) or non-match (FALSE). This plot shows several key insights about the LIME explanations. First, the features selected by LIME for the density simulation methods are mostly the same for all tuning parameters and all cases. Second, the features selected by LIME for the bin based methods are affected by the number of bins but not the Gower power. Third, the explanations produced by LIME using bin based methods result in different important features for the matches and non-matches. \kgc{***Need to rotate the facet labels, so I can make the text bigger.}}
\label{fig:bullet-feature-heatmap}

<<bullet-assess-metric, out.width = '\\textwidth', warning = FALSE, fig.width = 10, fig.height = 4>>=
